Empirical Bayes analysis of RNA-seq data for detection of gene expression heterosis

### 2015-03-03 ebayes_rnaseq_heterosis.tex
fixed typos

Following \cite{ji2014estimation}, we reparameterize the gene-variety mean structure into the gene-specific parental average ($\phi_g$), half-parental difference ($\alpha_g$), and hybrid difference ($\delta_g$). Specifically, if $V=3$ with $v=1,2$ indicate the two parental varieties and $v=3$ indicates the hybrid, then
\[ \phi_g = \frac{\mu_{g1}+\mu_{g2}}{2} \qquad \alpha_g = \frac{\mu_{g1}-\mu_{g2}}{2} \qquad \delta_g = \phi_g-\mu_{g3}. \]

We present a hierarchical model for the normalization factors, overdispersions, and gene-specific mean parameters. These normalization factors, parental averages, and overdispersion parameters are assumed to follow a normal distribution. 
\[ 
\gamma_{vi} \stackrel{ind}{\sim} N(0,\sigma_\gamma^2) \qquad 
\phi_g \stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2) \qquad 
\psi_g \stackrel{ind}{\sim} N(0,\sigma_\psi^2) \qquad 
\]
To allow for more shrinkage with the half-parental difference and hybrid difference, we use a Laplace (or double exponential prior) and thus implement a Bayesian LASSO \citep{park2008bayesian}. Thus
\[ 
\alpha_g \stackrel{ind}{\sim} La(\eta_\alpha, \sigma_\alpha) \qquad
\delta_g \stackrel{ind}{\sim} La(\eta_\delta, \sigma_\delta) 
\]
where $\alpha\sim La(\eta,\sigma)$ has a probability density function given by $p(\alpha|\eta,\sigma) = exp(-|\alpha-\eta|/\sigma)/2\sigma$, expectation $\eta$ and variance $2\sigma^2$. We assume independence amongst the normalization factors, parental averages, half parental differences, hybrid differences, and overdispersion parameters.

We assume improper uniform priors for $\eta_\phi$, $\eta_alpha$, and $\eta_\delta$ and independent half-cauchy priors for all standard deviations.

@@

We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$ and hyperparameters $\pi = (\gamma, \eta, \sigma)$, where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, $\eta=(\eta_\phi,\eta_alpha,\eta_delta)$, and $\sigma=(\sigma_gamma,\sigma_\phi,\sigma_\alpha,\sigma_delta,\sigma_psi)$. We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\pi$, denoted $\hat{\pi}$, and then base gene-specific inference on the posterior conditional on this MAP estimate. Conditional on the hyperparameters, the gene-specific parameters are independent, i.e., 
\begin{align}
p\left(\theta\left|y,\hat{\pi}\right.\right) 
%&= \prod_{g=1}^G p\left(\theta_g\left|y_g,\hat{\pi}\right.\right) = \prod_{g=1}^G p\left(y_g\left|\theta_g,\hat{\gamma}\right.\right) p\left(\theta_g\left|\hat{\pi}\right.\right) \nonumber \\
&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right.\right) \right] N\left(\left.\left[\mu_g, \psi_g \right]^\top\right|\hat{\eta}, \hat{\Sigma} \right). 
\label{e:condind}
\end{align}
Thus, conditional posterior inference for the gene-specific parameters can be obtained independently and in parallel.

>>>>

Following \cite{ji2014estimation}, we reparameterize the gene-variety mean structure into the gene-specific parental average ($\phi_g$), half-parental difference ($\alpha_g$), and hybrid difference ($\delta_g$). Specifically, for heterosis $V=3$ and let $v=1,2$ indicate the two parental varieties and $v=3$ indicate the hybrid, then
\[ \phi_g = \frac{\mu_{g1}+\mu_{g2}}{2}, \quad 
\alpha_g = \frac{\mu_{g1}-\mu_{g2}}{2}, \quad \mbox{and} \quad
\delta_g = \phi_g-\mu_{g3}. \]

We present a hierarchical model for the normalization factors, overdispersions, and gene-specific mean parameters. The normalization factors, parental averages, and overdispersion parameters are assumed to follow a normal distribution, i.e.
\[ 
\gamma_{vi} \stackrel{ind}{\sim} N(0,\sigma_\gamma^2), \quad 
\phi_g \stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2), \quad \mbox{and}\quad 
\psi_g \stackrel{ind}{\sim} N(0,\sigma_\psi^2). 
\]
To allow for more shrinkage with the half-parental and hybrid differences, we use a Laplace (or double exponential prior) and thus implement a Bayesian LASSO \citep{park2008bayesian}, i.e. 
\[ 
\alpha_g \stackrel{ind}{\sim} La(\eta_\alpha, \sigma_\alpha) \qquad \mbox{and} \qquad
\delta_g \stackrel{ind}{\sim} La(\eta_\delta, \sigma_\delta) 
\]
where $\alpha\sim La(\eta,\sigma)$ has a probability density function given by $p(\alpha|\eta,\sigma) = \exp(-|\alpha-\eta|/\sigma)/2\sigma$, expectation $\eta$, and variance $2\sigma^2$. We assume independence amongst the normalization factors, parental averages, half parental differences, hybrid differences, and overdispersion parameters.

We assume improper uniform priors for $\eta_\phi$, $\eta_\alpha$, and $\eta_\delta$ and independent half-cauchy priors for all standard deviations \citep{gelman2006prior}.

@@

We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ where $\theta_g = (\phi_g,\alpha_g,\delta_g,\psi_g)$ and hyperparameters $\pi = (\gamma, \eta, \sigma)$, where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$, $\eta=(\eta_\phi,\eta_\alpha,\eta_\delta)$, and $\sigma=(\sigma_\gamma,\sigma_\phi,\sigma_\alpha,\sigma_\delta,\sigma_psi)$. We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\pi$, denoted $\hat{\pi}$, and then base gene-specific inference on the posterior conditional on this MAP estimate. Conditional on the hyperparameters, the gene-specific parameters are independent, i.e., 
\begin{align}
p\left(\theta\left|y,\hat{\pi}\right.\right) 
%&= \prod_{g=1}^G p\left(\theta_g\left|y_g,\hat{\pi}\right.\right) = \prod_{g=1}^G p\left(y_g\left|\theta_g,\hat{\gamma}\right.\right) p\left(\theta_g\left|\hat{\pi}\right.\right) \nonumber \\
&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right.\right) \right] N\left(\left.\left[\mu_g, \psi_g \right]^\top\right|\hat{\eta}, \hat{\Sigma} \right). 
\label{e:condind}
\end{align}
Thus, conditional posterior inference for the gene-specific parameters can be obtained independently and in parallel.
